Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Mechanical Perturbations

### 2024-09-14 main.tex
Fix force in caption.
Use biblatex.
Editing in methods.

\usepackage{amsmath}
\usepackage{booktabs}  % nice tables
\usepackage[margin=25mm]{geometry}
\usepackage{natbib}
\usepackage{siunitx}  % use for all units
\usepackage{subcaption}  % for subfigures
\usepackage{todonotes}

@@

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll rate feedback
controller to stabilize the bicycle. In the speed range tested, the commanded
steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the control
law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = g(v_{stable} - v)\dot{\phi}
\end{align}
%
where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
average stable speed predicted from the open loop bicycle rigid-rider dynamics.
We use \(g=8\) and \(g=10\) as gain values during the experiments. Scaling the
proportional feedback gain linearly with respect to speed stabilizes the
normally unstable weave mode of the bicycle down to
2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
4.6~\si{\kilo\meter\per\hour} for the ridden bicycle  as shown in
Figure~\ref{fig:eigenvalues}~\todo{Could add the experimental weave mode
identification to this figure and consider lowering the gain on the model to
make the model match the experimental values.}.
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row) and controlled (lower row) root locus of
    eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
    for the bicycle without the inertial effect of a rigid rider (left) and
    with a rigid rider (right). Vertical blue lines indicate the two speeds we
    perform experiments at. Grey shaded region is the linear stable speed
    range. Geometric and inertial parameters for these plots were estimated
    with the methods presented in \cite{Moore2012} and software packages
    BicycleParameters 1.1.1~\citep{Moore2024} and Yeadon 1.5.0~\citep{Dembia2015}.}
  \label{fig:eigenvalues}
\end{figure}

\subsection{Perturbations}
%
We apply external longitudinal forces to the ends of each handlebar using an
adapted Bump'Em system~\cite{Tan2020} which is designed with four motors
working in cooperation, Figure~\ref{fig:setup-diagram}. The four motors are
programmed to apply a light force at all times to keep the ropes taught and to
track a commanded force profile using a PID controller running on a
microprocessor (Arduino Mega 2560, Arduino, Italy). We measure the force
applied by each motor at the handlebar via four inline 250~\si{\newton} load
cells. The left rear load cell was damaged during pilot experiments, and is
replaced by a load cell with a max load of 500~\si{\newton}. The commanded
force profiles are designed to apply an external pulsive torque to the front
assembly (handlebars, forks, wheel) at magnitudes varying from 16 to
160~\si{\newton\meter}. The four motors are arranged at the four corners of a
1~\si{\meter} wide treadmill (EC-90 Flat, Maxon Group, Switzerland) that can
reach speeds of 18~\si{\kilo\meter\per\hour}. The general design of this system
is described in detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and
the physical arrangement is shown in Figure~\ref{fig:participant-in-set-up}.
Our modifications relative to Van De Velde include simplifying the controller
with a inexpensive microcontroller and the use of a simpler non-active safety
harness.
%
% NOTE : I get a compile error without the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[width=75mm]{figures/participant-in-set-up.png}}
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[width=65mm]{figures/setup-diagram.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations.}
  \label{fig:experimental-setup}
\end{figure}

\subsection{Protocol}
%
We recruited 26 able-bodied young adults (20-36 years old) to participate in
the experiments. The subjects all were confident in their cycling skills and
had cycled at least once in the last month. Eleven subjects performed the
experiments at 1.7~\si{\meter\per\second} (6~\si{\kilo\meter\per\hour}) and
fifteen subjects performed the experiments at 2.8~\si{\meter\per\second}
(10~\si{\kilo\meter\per\hour}). All subjects consented to the experiment and
could decline to continue at any time. The study was approved by Delft
University of Technology's Human Research Ethics Council (\#3897).

@@

The subjects were divided into two groups. The first group performed the
protocol at a belt speed of 2.8~\si{\meter\per\second}
(10~\si{\kilo\meter\per\hour}) with the controller gain at $g=8$ and the second
group performed the protocol at a belt speed of 1.7~\si{\meter\per\second}
(6~\si{\kilo\meter\per\hour}). The 10~\si{\kph} group's experiments occurred
two weeks before the 6~\si{\kilo\meter\per\hour} group.

Subjects wore a helmet and fall safety harness attached to the ceiling
Figure~\ref{fig:participant-in-set-up}. We allowed them to practice riding on
the treadmill until they indicated they were comfortable enough to have
perturbations applied. For most, this was less than a 10~\si{\minute} warm up.
We then asked the rider to ride for 90~\si{\second}, while attempting to
maintain the location of their front wheel on the center line of the treadmill.
We define a ``fall'' on the treadmill by two criteria: 1) the rider removes
their foot from the pedal and places it on the ground or 2) the bicycle wheel
exceeds the width of the treadmill belt. We then applied perturbations in
random directions (clockwise or counter-clockwise), starting at 20~\si{\newton}
and increasing the magnitude by 30~\si{\newton\meter} until the participants
fell. Figure~\ref{fig:perturbation-sequence} shows an example resulting motion
from a perturbation. We log the magnitude that causes the first fall to
characterize that subject's nominal fall threshold.
%
\begin{figure}
  \centering
  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

Following the initial threshold determination, we choose perturbation forces
according to a random and adaptive staircase procedure~\cite{TODO} applying
perturbations above and below the initial perturbation threshold, while
allowing small progression of the perturbation threshold to accommodate
learning effects. Five possible perturbation forces are determined based on the
initial perturbation threshold estimate: the initial estimate itself, two
perturbations lower than the initial estimate and two perturbations higher than
the initial estimate.  The five possible perturbations are separated by 10 N
steps. For example, if the initial estimate of the perturbation threshold of a
participant is 80~\si{\newton}, the five possible perturbations are 60, 70, 80,
90 and 100~\si{\newton}. Which one of these five perturbations is chosen is
determined at random. If the perturbation results in a fall, the estimate of
the perturbation threshold is decreased with 10~\si{\newton}, and vice versa if
the perturbation did not result in a fall. Five new possible perturbation
forces are determined around the updated perturbation threshold, and a new
perturbation is selected at random. This process iterates until twenty
perturbations are applied. The goal of this adaptive staircase procedure is to
have participants fall for approximately 50\% of the time. We apply 20
perturbations in this manner while the subject rides at a constant speed and
record which perturbations cause a fall. We let the cyclist rest and then
perform another 20 perturbations. We randomize whether the balance assist
system is on or off during the first or second set of perturbations.

\subsection{Measurements}
%
We measure the time histories of the Bump'Em delivered perturbation forces and
the bicycle's steer angle, roll angle, roll angular rate, and forward speed.
Figure~\ref{fig:perturbation_0} shows an example perturbation force measurement
compared to our Bump'Em controller command.
%
\begin{figure}
  \centering
  \includegraphics[width=140mm]{figures/torque_angle_perturbation_10.png}
  \caption{Motor force and resulting  steer torque perturbation based on an 80N
    counterclockwise applied force.}
  \label{fig:perturbation_0}
\end{figure}

Based on findings from measuring riders without balance
assist~\cite{Haitjema2023a}, we calculate several variables that we hypothesize
may influence fall probability. We use the angular impulse \(L\) of the
perturbation forces over a 0.3~\si{\second} duration to characterize the
magnitude of delivered perturbation. The duration is selected based on the
duration of the commanded force and is calculated as follows:
%
\begin{align}
  L =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}(F_r + F_l) dt
  =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left[(F_{rf} - F_{rr}) + (F_{lf}
  - F_{lr})\right] dt
  \textrm{.}
\end{align}

At the initiation of each perturbation we log the steer and roll angles to
characterize the configuration of the bicycle when perturbed. The gain setting
on the balance assist controller indicates if the assistance is off \(g=0\) or
on at two different levels: low \(g=8\) or high \(g=10\). A recovery from the
perturbation is successful if the person neither put their foot down onto the
treadmill surface or either wheel of the bicycle exits the width of the
treadmill belt. We record this as a binary variable \(f\). All measured
variables are reported in Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall outcome'' and
    ``Perturbation order number'' are recorded per perturbation instance.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain & \(k\) & TODO & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
    Fall outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear & \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) &  \si{\degree} & Kalman estimate~\citep{Gabriel2023} \\
    Roll Angular Rate & \(\dot{\phi}\) &  \si{\degree\per\second} & rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

\subsection{Statistics}
%
We test the hypothesis that the balance assistance controller will reduce the
probability of falling when perturbed externally at the handlebar. We have a
single binary fall outcome variable \(f\) that is dependent on several possible
explanatory independent variables, one of which is the binary balance
assistance (on or off). See Table~\ref{tab:stat-model-variables} for all of the
variables.
%
\begin{table}
  \centering
  \caption{Statistical model variables.}
  \label{tab:stat-model-variables}
  \begin{tabular}{lll}
    \toprule
    Variable & Units & Description \\
    \midrule
    \(L\) & \si{\newton\meter\second} & angular impulse of perturbation torque \\
    \(c\) & integer & order number of perturbation \\
    \(k\) & TODO & gain of balance assist control \\
    \(\delta_0\) & \si{\degree} & steer angle at start of perturbation \\
    \(\phi_0\) & \si{\degree} & roll angle at start of perturbation \\
    \(v\) & \si{\meter\per\second} & forward speed (1.7 or 2.8) \\
    \(f\) & boolean & outcome of perturbation: did not fall, did fall \\
    \bottomrule
  \end{tabular}
\end{table}

@@

Turning the balance-assist system on significantly \((p<0.05)\) reduces the
odds that a pertubation results in a fall occurs while cycling at a speed of
6~\si{\kph}.  Figure~\ref{fig:probability-6kph} visualizes the probability of
falling as a function of the mean and centered angular impulse per participant
while keeping all other explantory factors at their centered mean values. This
figure is derived by setting all explanatory variables to there mean and
calculating the probality from Eq~\ref{eq:log-regress} for only change in
angular impulse given the estimates in Table~\ref{tab:freq-coeffs-6} which
indicate that the balance-assist system halves (0.53) the odds that a
perturbation results in a fall. This figure shows that for relatively large
impulses the probability to fall is unity for both states of balance assist on
and off. And for relatively small impulses the probability to fall is zero for
both states. But for impulses in the magnitude region (-1 to 0.5 STD) around
the mean-centered fall threshold, the probability of falling is significantly
lowered. The skewedness of the probaility curves arrives from the interaction
effects\todo[inline]{check this and think about it}.

@@

\section{Discussion}
%
The probability that a fall occurs depends on the values of all the independent
variables in Table~\ref{tab:stat-model-variables}, but we can visualize the
effect of one or two variables (e.g. Figure~\ref{fig:probability}) to gain
some insight.
But to interpret the results in Tables~\ref{tab:freq-coeffs-6} and
\ref{tab:freq-coeffs-10} it is important to understand the relationship between
probability and odds.
The estimate in Table~\ref{tab-freq-coeffs-6} shows that the balance assist
system halves the odds that a perturbation results in a fall
(MCiO=0.53)~\todo{Make a variable for MCiO}.
This means if the odds are a 1000:1, turning on the balance-assist system
reduces the odds to 500:1.
However, the probability that a fall occurs is only reduced from $0.999$ to
$0.998$ in that case.
If the odds that a fall will occur are smaller, halving the odds has a larger
influence on the fall probability.
For example, if the odds that a fall occurs is two, halving it to one reduces
the probability from $0.66$ to $0.50$.
This can be seen in Figure~\ref{fig:probablity} which shows the fall
probability as a function of the magnitude of the normalized angular impulse
for both speeds.
The difference in probability between the balance-assist on/off case becomes
insignificant outside of approximately \(\pm1\) standard deviation of the
average angular impulse.
This means that the balance assist is most effective for perturbations close to
the subject's personal threshold between falling or recovery and that large
perturbations will make you fall regardless of the balance assist's help.

@@

Angular impulse magnitude has the largest significant effect for predicting
fall probability, Tables~\ref{tab:freq-coeffs-6} and \ref{tab:freq-coeffs-10}.
An increase in angular impulse increases the fall probability both at 6 and
10~\si{\kph}. At 10~\si{\kph}, the multiplicative change in odds is
approximately twice as big as at 6~\si{\kph}. Thus, angular impulse is a more
important predictor at higher speeds compared to lower speeds. The reason for
this likely has to do with the width of the treadmill and may also be why
balance assist system did not have a statistically significant effect at
10~\si{\kph}. As a bicycle travels at higher speeds, the same perturbation
causes larger lateral deviations. At 10~\si{\kph} almost all falls were due to
the bicycle exiting the maximum width of the treadmill. If the same experiment
was performed on an infinite plane, the riders may have recovered from more
perturbations. At 6~\si{\kph} the riders could often recover in the allotted
treadmill width. Our results are very much dependent on the two modes of
falling with use: exit the treadmill width or foot is placed on the belt. Cycle
paths are a similar width as the treadmill, so rider's are often limited in
width when recovering from a fall.

@@

\bibliographystyle{apalike}
% NOTE : this file is automatically generated from Zotero, do not edit
% manually!
\bibliography{references}

>>>>

\usepackage{amsmath}
\usepackage{booktabs}  % nice tables
\usepackage[margin=25mm]{geometry}
\usepackage[natbib=true,style=authoryear]{biblatex}
\addbibresource{references.bib}
\usepackage{siunitx}  % use for all units
\usepackage{subcaption}  % for subfigures
\usepackage{todonotes}

@@

\subsection{Balance Assist Control}
%
We use a forward speed \(v\) gain scheduled proportional roll rate feedback
controller to stabilize the bicycle. In the speed range tested, the commanded
steer torque \(T^\textrm{m}_\delta\) from the steer motor follows the control
law
%
\begin{align}
  T^\textrm{m}_\delta = -k_{\dot{\phi}}\dot{\phi} = g(v_{stable} - v)\dot{\phi}
\end{align}
%
where \(v_{\textrm{stable}} = 4.7~\si{\meter\per\second}\) is approximately the
average stable speed predicted from the open loop bicycle rigid-rider dynamics.
We use \(g=8\) and \(g=10\) as gain values during the experiments. Scaling the
proportional feedback gain linearly with respect to speed stabilizes the
normally unstable weave mode of the bicycle down to
2.9~\si{\kilo\meter\per\hour} for the riderless bicycle and
4.6~\si{\kilo\meter\per\hour} for the ridden bicycle  as shown in
Figure~\ref{fig:eigenvalues}.
\todo[inline]{Could add the experimental weave mode identification to this
figure and consider lowering the gain on the model to make the model match the
experimental values.}
%
\begin{figure}
  \centering
  \includegraphics[width=160mm]{figures/balance-assist-eig-vs-speeds.png}
  \caption{Uncontrolled (upper row) and controlled (lower row) root locus of
    eigenvalue components (real: solid, imaginary: dashed) plotted versus speed
    for the bicycle without the inertial effect of a rigid rider (left) and
    with a rigid rider (right). Vertical blue lines indicate the two speeds we
    perform experiments at. Grey shaded region is the linear stable speed
    range. Geometric and inertial parameters for these plots were estimated
    with the methods presented in \cite{Moore2012} and software packages
    BicycleParameters 1.1.1~\citep{Moore2024} and Yeadon 1.5.0~\citep{Dembia2015}.}
  \label{fig:eigenvalues}
\end{figure}

\subsection{Perturbations}
%
We apply longitudinal forces to the ends of each handlebar from Earth fixed
posts using an adapted Bump'Em system~\citep{Tan2020} which we arrange with
four motors working in cooperation, Figure~\ref{fig:setup-diagram}. The four
motors are programmed to apply a light force at all times to keep the ropes
taught and to track a commanded force profile using a PID controller running on
a microprocessor (Arduino Mega 2560, Arduino, Italy). We measure the force
applied by each motor at the handlebar via four inline 250~\si{\newton} S-style
load cells except the left rear load cell that has a max load of
500~\si{\newton}. The commanded force profiles are designed to apply an
external pulsive torque to the front assembly (handlebars, forks, wheel) at
magnitudes varying from \SIrange{16}{160}{\newton\meter}. The four motors
(EC-90 Flat, Maxon Group, Switzerland) are arranged at the four corners of a
1~\si{\meter} wide treadmill that can reach speeds of
18~\si{\kilo\meter\per\hour}. The general base design of this system is
described in detail in Van De Velde's MSc thesis~\citep{vandeVelde2022a} and
the physical arrangement is shown in Figure~\ref{fig:participant-in-set-up}.
Our modifications relative to Van De Velde include simplifying the controller
with a inexpensive microcontroller and the use of a simpler non-active safety
harness.
%
% NOTE : I get a compile error without the \protect calls here, not sure why.
\begin{figure}
  \centering
  \subcaptionbox{A participant on the bicycle in the safety harness with the
    Bump'Em motor lines attached to the ends of the
    handlebars.\protect\label{fig:participant-in-set-up}}
    {\includegraphics[width=75mm]{figures/participant-in-set-up.png}}
  \subcaptionbox{Top view diagram of bicycle handlebar perturbation
    system.\protect\label{fig:setup-diagram}}
    {\includegraphics[width=65mm]{figures/setup-diagram.png}}
  \caption{Diagrams showing the Bump'Em system arranged to apply handlebar
    perturbations.}
  \label{fig:experimental-setup}
\end{figure}

\subsection{Protocol}
%
We recruited 26 able-bodied young adults (20-36 years old) to participate in
the experiments. The subjects were all confident in their cycling skills and
had cycled at least once in the last month. Eleven subjects performed the
experiments at 6~\si{\kilo\meter\per\hour} (1.7~\si{\meter\per\second}) and
fifteen subjects performed the experiments at 10~\si{\kilo\meter\per\hour}
(2.8~\si{\meter\per\second}). All subjects consented to the experiment and
could decline to continue at any time. The study was approved by Delft
University of Technology's Human Research Ethics Council (\#3897).

@@

The subjects were divided into two groups. The first group performed the
protocol at a belt speed of 10~\si{\kilo\meter\per\hour} with the controller
gain set to $g=8$ and the second group performed the protocol at a belt speed
of 6~\si{\kilo\meter\per\hour} with the controller gain set to $g=10$. The
10~\si{\kph} group's experiments occurred two weeks before the
6~\si{\kilo\meter\per\hour} group.

Subjects wore a helmet and fall safety harness attached to the ceiling
Figure~\ref{fig:participant-in-set-up}. We allowed them to practice riding on
the treadmill until they indicated they were comfortable enough to have
perturbations applied. For most, this was less than a 10~\si{\minute} warm up.
We then asked the rider to ride for 90~\si{\second}, while attempting to
maintain the location of their front wheel on the center line of the treadmill.
We define a ``fall'' on the treadmill by two criteria: 1) the rider removes
their foot from the pedal and places it on the ground or 2) the bicycle wheel
exceeds the width of the treadmill belt. We then applied perturbations in
random directions (clockwise or counter-clockwise), starting at 20~\si{\newton}
and increasing the magnitude by 30~\si{\newton} until the participants
fell. Figure~\ref{fig:perturbation-sequence} shows an example resulting motion
from a perturbation. We log the magnitude that causes the first fall to
characterize that subject's \emph{nominal fall threshold}.
%
\begin{figure}
  \centering
  \subcaptionbox{Before perturbation}{\includegraphics[width=60mm]{figures/start.png}}
  \subcaptionbox{During perturbation}{\includegraphics[width=60mm]{figures/during.png}}
  \subcaptionbox{After perturbation}{\includegraphics[width=60mm]{figures/after.png}}
  \subcaptionbox{Recovery from perturbation}{\includegraphics[width=60mm]{figures/recovery.png}}
  \caption{Video frames depicting the application of a perturbation and the
  rider's response and recovery.}
  \label{fig:perturbation-sequence}
\end{figure}

Following the initial threshold determination, we choose perturbation forces
according to a random and adaptive staircase procedure applying perturbations
above and below the initial perturbation threshold, while allowing small
progression of the perturbation threshold to accommodate learning effects. Five
possible perturbation forces are determined based on the initial perturbation
threshold estimate: the initial estimate itself, two perturbations lower than
the initial estimate, and two perturbations higher than the initial estimate.
The five possible perturbations are separated by 10~\si{\newton} steps. For
example, if the initial estimate of the perturbation threshold of a participant
is 80~\si{\newton}, the five possible perturbations are 60, 70, 80, 90 and
100~\si{\newton}. Which one of these five perturbations is chosen is determined
at random. If the perturbation results in a fall, the estimate of the
perturbation threshold is decreased by 10~\si{\newton}, and vice versa if the
perturbation did not result in a fall. Five new possible perturbation forces
are determined around the updated perturbation threshold, and a new
perturbation is selected at random. This process iterates until twenty
perturbations are applied. The goal of this adaptive staircase procedure is to
have participants fall for approximately 50\% of the time. After the first set
of 20 perturbations, we let the cyclist rest and then perform another 20
perturbations. We randomize whether the balance assist system is on or off
during the first or second set of perturbations for each participant.

\subsection{Measurements}
%
We measure the time histories of the Bump'Em delivered perturbation forces and
the bicycle's steer angle, roll angle, roll angular rate, and forward speed.
Figure~\ref{fig:perturbation_0} shows an example perturbation force measurement
compared to our Bump'Em controller command.
%
\begin{figure}
  \centering
  \includegraphics[width=140mm]{figures/torque_angle_perturbation_10.png}
  \caption{Motor force and resulting  steer torque perturbation based on an 110N
    counterclockwise applied force.}
  \label{fig:perturbation_0}
\end{figure}

Based on findings from measuring riders without balance
assist~\cite{Haitjema2023a}, we calculate several variables that we hypothesize
may influence fall probability. We use the angular impulse \(L\) of the
perturbation forces over a 0.3~\si{\second} duration to characterize the
magnitude of delivered perturbation. The duration is selected based on the
duration of the commanded force and is calculated as follows:
%
\begin{align}
  L =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}(F_r + F_l) dt
  =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left[(F_{rf} - F_{rr}) + (F_{lf}
  - F_{lr})\right] dt
  \textrm{.}
  \label{eq:angular-impulse}
\end{align}

where \(F_r\) and \(F_l\)  is the total force applied on the right and left
handlebar ends, respectively which are the sum of the rear and front load cell
readings \(F_{rr}\) and \(F_{rf}\), for example. The handlebar length is given
as \(l\) in Equation \ref{eq:angular-impulse}.

At the initiation of each perturbation we log the instananeous steer and roll
angles to characterize the configuration of the bicycle when perturbed. The
gain setting on the balance assist controller indicates if the assistance is
off \(g=0\) or on at two different levels: low \(g=8\) or high \(g=10\). A
recovery from the perturbation is successful if the person neither places their
foot down onto the treadmill surface or their wheel of the bicycle exits the
width of the treadmill belt. We record this as a binary variable \(f\) for
"fall". All measured variables are reported in Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall outcome'' and
    ``Perturbation order number'' are recorded per perturbation instance.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain & \(g\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
    Fall outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear & \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

\subsection{Statistics}
%
We test the hypothesis that the balance assistance controller will reduce the
probability of falling when perturbed externally at the handlebar. We have a
single binary fall outcome variable \(f\) that is dependent on several possible
explanatory independent variables, one of which is the binary balance
assistance (on or off). See Table~\ref{tab:stat-model-variables} for all of the
variables.
%
\begin{table}
  \centering
  \caption{Statistical model variables.}
  \label{tab:stat-model-variables}
  \begin{tabular}{lll}
    \toprule
    Variable & Units & Description \\
    \midrule
    \(L\) & \si{\newton\meter\second} & angular impulse of perturbation torque \\
    \(c\) & integer & order number of perturbation \\
    \(g\) & \si{\newton\second\squared} & gain of balance assist control \\
    \(\delta_0\) & \si{\degree} & steer angle at start of perturbation \\
    \(\phi_0\) & \si{\degree} & roll angle at start of perturbation \\
    \(v\) & \si{\meter\per\second} & forward speed (1.7 or 2.8) \\
    \(f\) & boolean & outcome of perturbation: did not fall, did fall \\
    \bottomrule
  \end{tabular}
\end{table}

@@

Turning the balance-assist system on significantly \((p<0.05)\) reduces the
odds that a pertubation results in a fall occurs while cycling at a speed of
6~\si{\kph}.  Figure~\ref{fig:probability-6kph} visualizes the probability of
falling as a function of the mean and centered angular impulse per participant
while keeping all other explantory factors at their centered mean values. This
figure is derived by setting all explanatory variables to there mean and
calculating the probality from Eq~\ref{eq:log-regress} for only change in
angular impulse given the estimates in Table~\ref{tab:freq-coefs-6} which
indicate that the balance-assist system halves (0.53) the odds that a
perturbation results in a fall. This figure shows that for relatively large
impulses the probability to fall is unity for both states of balance assist on
and off. And for relatively small impulses the probability to fall is zero for
both states. But for impulses in the magnitude region (-1 to 0.5 STD) around
the mean-centered fall threshold, the probability of falling is significantly
lowered. The skewedness of the probaility curves arrives from the interaction
effects\todo[inline]{check this and think about it}.

@@

\section{Discussion}
%
The probability that a fall occurs depends on the values of all the independent
variables in Table~\ref{tab:stat-model-variables}, but we can visualize the
effect of one or two variables (e.g. Figure~\ref{fig:probability}) to gain
some insight.
But to interpret the results in Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10} it is important to understand the relationship between
probability and odds.
The estimate in Table~\ref{tab:freq-coefs-6} shows that the balance assist
system halves the odds that a perturbation results in a fall
(MCiO=0.53)~\todo{Make a variable for MCiO}.
This means if the odds are a 1000:1, turning on the balance-assist system
reduces the odds to 500:1.
However, the probability that a fall occurs is only reduced from $0.999$ to
$0.998$ in that case.
If the odds that a fall will occur are smaller, halving the odds has a larger
influence on the fall probability.
For example, if the odds that a fall occurs is two, halving it to one reduces
the probability from $0.66$ to $0.50$.
This can be seen in Figure~\ref{fig:probability} which shows the fall
probability as a function of the magnitude of the normalized angular impulse
for both speeds.
The difference in probability between the balance-assist on/off case becomes
insignificant outside of approximately \(\pm1\) standard deviation of the
average angular impulse.
This means that the balance assist is most effective for perturbations close to
the subject's personal threshold between falling or recovery and that large
perturbations will make you fall regardless of the balance assist's help.

@@

Angular impulse magnitude has the largest significant effect for predicting
fall probability, Tables~\ref{tab:freq-coefs-6} and \ref{tab:freq-coefs-10}.
An increase in angular impulse increases the fall probability both at 6 and
10~\si{\kph}. At 10~\si{\kph}, the multiplicative change in odds is
approximately twice as big as at 6~\si{\kph}. Thus, angular impulse is a more
important predictor at higher speeds compared to lower speeds. The reason for
this likely has to do with the width of the treadmill and may also be why
balance assist system did not have a statistically significant effect at
10~\si{\kph}. As a bicycle travels at higher speeds, the same perturbation
causes larger lateral deviations. At 10~\si{\kph} almost all falls were due to
the bicycle exiting the maximum width of the treadmill. If the same experiment
was performed on an infinite plane, the riders may have recovered from more
perturbations. At 6~\si{\kph} the riders could often recover in the allotted
treadmill width. Our results are very much dependent on the two modes of
falling with use: exit the treadmill width or foot is placed on the belt. Cycle
paths are a similar width as the treadmill, so rider's are often limited in
width when recovering from a fall.

@@

% NOTE : this file is automatically generated from Zotero, do not edit
% manually!
\printbibliography
